\LARGE{\bf{Natural variation in teosinte at the domestication locus \emph{teosinte branched1} (\emph{tb1})}}\footnotemark

### 2014-11-05 manuscript.tex
changed fig names, added figs back in for biorxiv submission
Edits
Edits to manuscript for future submission to PeerJ

\newcommand{\jri}[1]{\textcolor{blue}{ \emph{\scriptsize  #1}} } %introducting new command "\jri" for jeff ross-ibarra to make comments in blue italics
\newcommand{\mbh}[1]{\textcolor{red}{ \emph{\scriptsize  #1}} } %same for matt hufford in red
\newcommand{\lev}[1]{\textcolor{green}{\emph{\scriptsize #1}} } %same for laura vann in green

@@

\emph{Methods:} Prevalence of the \emph{Hopscotch} element was assessed across an Americas-wide sample of 837 maize and teosinte individuals using a co-dominant PCR assay.  Population genetic summaries were calculated for a subset of individuals from four teosinte populations in central Mexico. Phenotypic data were also collected using seed from a single teosinte population where \emph{Hopscotch} was found segregating at high frequency.

@@

\textbf{Sampling and genotyping---}We sampled 1,110 individuals from 350 accessions (247 maize landraces, 17 \emph{mexicana} populations, and 86 \emph{parviglumis} populations; ranging from 1-38 individuals per population, with an average of 11 individuals per population for \emph{parviglumis} and \emph{mexicana} and 2 individuals per landrace accession) and assessed the presence or absence of the \emph{Hopscotch} insertion (Appendix 1 and Appendix 2, See Supplemental Materials with the online version of this article). DNA was extracted from leaf tissue using a modified CTAB approach \citep{DoyleDoyle1990, Maloof1984}. We designed primers using PRIMER3 \citep{RozenSkaletsky2000} implemented in Geneious \citep{Kearse2012} to amplify the entire \emph{Hopscotch} element, as well as an internal primer allowing us to simultaneously check for possible PCR bias between presence and absence of the \emph{Hopscotch} insertion due to its large size ($\sim$5kb). Two PCRs were performed for each individual, one with primers flanking the \emph{Hopscotch} (HopF/HopR) and one with a flanking primer and an internal primer (HopF/HopIntR). Primer sequences are HopF, {\small 5'-TCGTTGATGCTTTGATGGATGG-3'}; 
HopR, {\small 5'-AACAGTATGATTTCATGGGACCG-3'}; and HopIntR, {\small  5'-CCTCCACCTCTCATGAGATCC-3'} (Appendix 3 and Appendix 4, See Supplemental Materials with the online version of this article). Homozygotes show a single band for absence of the element ($\sim$300bp) and two bands for presence of the element ($\sim$5kb, amplification of the entire element, and $\sim$1.1kb, amplification of part of the element), whereas heterozygotes show all three bands (Appendix 2, See Supplemental Materials with the online version of this article). Since we developed a PCR protocol for each allele, if only one PCR resolved well, we scored one allele for that individual rather than infer the diploid genotype. We used Phusion High Fidelity Enzyme (Thermo Fisher Scientific Inc., Waltham, Massachusetts, USA) and the following conditions for amplifications: 98$^{\circ}$C for 3 min, 30 cycles of 98$^{\circ}$C for 15 s, 65$^{\circ}$C for 30 s, and 72$^{\circ}$C for 3 min 30 s, with a final extension of 72$^{\circ}$C for 10 min. PCR products were visualized on a 1\% agarose gel and scored for presence/absence of the \emph{Hopscotch} based on band size.

@@

We examined evidence of introgression on chromosome 1 in these same four populations (EjuA, EjuB, MSA, SLO) using STRUCTURE \citep{Falush2003} and phased data from \citet{Pyhajarvi2013}, combined with the corresponding SNP data from a diverse panel of 282 maize lines \citep{Cook2012}. SNPs were anchored in a modified version of the IBM genetic map \citep{Gerke2013}. Since STRUCTURE does not account for LD due to physical linkage we created haplotype blocks using a custom Perl script from \citet[][code available at \url{http://dx.doi.org/10.6084/m9.figshare.1165577}]{Hufford2013}.  In maize, LD decays over an average distance of 5500bp \citep{Chia2012}; because LD decay is even more rapid in teosinte \citep{Chia2012} we used a conservative haplotype block size of 5kb.  We ran STRUCTURE at K=2 under the linkage model, performing 3 replicates with an MCMC burn-in of 10,000 steps and 50,000 steps post burn-in.

@@

\textbf{Genotyping---}The genotype at the \emph{Hopscotch} insertion was confirmed with two PCRs for 837 individuals of the 1,100 screened. Among the 247 maize landrace accessions genotyped, all but eight were homozygous for the presence of the insertion (Appendix 1 and Appendix 2, See Supplemental Materials with the online version of this article). Within our \emph{parviglumis} and \emph{mexicana} samples we found the \emph{Hopscotch} insertion segregating in 37 ($n=86$) and four ($n=17$) populations, respectively, and at highest frequency within populations in the states of Jalisco, Colima, and Michoac\'{a}n in central-western Mexico (Fig. \ref{Fig1Map}). Using our \emph{Hopscotch} genotyping, we calculated differentiation between populations (F\textsubscript{ST}) and subspecies (F\textsubscript{CT}) for populations in which we sampled eight or more individuals. We found that F\textsubscript{CT} = 0, and levels of F\textsubscript{ST} among populations within each subspecies (0.22) and among all populations (0.23) are similar to genome-wide estimates from previous studies (\citealt{Pyhajarvi2013}; Table \ref{Table1Fst}). Although we found large variation in \emph{Hopscotch} allele frequency among our populations, BayEnv analysis did not indicate a correlation between the \emph{Hopscotch} insertion and environmental variables (all Bayes Factors $<$ 1).

@@

%Figure 1 Map 
%--------------------------------------------
\begin{figure*}[!t]
\begin{center}
   %\includegraphics[width=0.6\textwidth]{Fig1.MapBalsas.pdf}
   \end{center}
    \caption{Map showing the frequency of the \emph{Hopscotch} allele in populations of \emph{parviglumis} where we sampled more than 6 individuals. Size of circles reflects number of alleles sampled. The Balsas River is shown, as the Balsas River Basin is believed to be the center of domestication of maize.} 
\label{Fig1Map}
\end{figure*}
%--------------------------------------------

% Figure 2. LD plots
%-------------------------------------------------------------------
\begin{figure*}[!t]
  \begin{center}
   %\includegraphics[width=140mm]{Fig2LDPlots.pdf}
   \end{center}
    \caption{Linkage disequilibrium for SNPs in Mb 261-268 on chromosome 1. The yellow rectangle indicates the location of the \emph{Hopscotch} insertion and the green represents the \emph{tb1} ORF. A) Ejutla A; B) Ejutla B; C) La Mesa; D) San Lorenzo. The upper triangle above the black diagonal is colored based on the $r^{2}$ value between SNPs while the bottom triangle is colored based on p-value for the corresponding  $r^{2}$ value.}
\label{Fig2LD}
\end{figure*}
%-------------------------------------------------------------------

% Figure 3 STRUCTURE Plot 4 Pops
%-------------------------------------------------------------------
\begin{figure*}[!t]
  \begin{center}
   %\includegraphics[width=90mm]{Fig3Structure.pdf}
     \end{center}
    \caption{STRUCTURE assignment to maize across a section of chromosome 1. The dotted lines mark the beginning of the sequenced region 66kb upstream (Region 2) and the end of the \emph{tb1} ORF.}
\label{Fig3Structure}
\end{figure*}
%-------------------------------------------------------------------

% Figure 4 Phenotyping BoxPlots
%-------------------------------------------------------------------
\begin{figure*}[!t]
  \begin{center}
   %\includegraphics[width=90mm]{Fig4.BoxplotPheno1.pdf}
   \end{center}
    \caption{Box-plots showing tillering index in greenhouse grow-outs for phenotyping. White indicates individuals homozygous for the \emph{Hopscotch}, light grey represents heterozygotes, and dark grey represents homozygotes for the teosinte (No \emph{Hopscotch}) allele. Within boxes, dark black lines represent the median, and the edges of the boxes are the first and third quartiles. Outliers are displayed as dots, the maximum value excluding outliers is shown with the top whisker, while the minimum excluding outliers is shown with the bottom whisker.}
\label{Fig4Boxplots}
\end{figure*}
%-------------------------------------------------------------------

>>>>

\newcommand{\jri}[1]{\textcolor{blue}{ \emph{\scriptsize  #1}} } %introducting new command "\jri" for jeff ross-ibarra to make comments in blue italics
\newcommand{\mbh}[1]{\textcolor{red}{ \emph{\scriptsize  #1}} } %same for matt hufford in red
\newcommand{\lev}[1]{\textcolor{magenta}{\emph{\normalsize #1}} } %same for laura vann in green

@@

\emph{Methods:} Prevalence of the \emph{Hopscotch} element was assessed across an Americas-wide sample of \lev{837} maize and teosinte individuals using a co-dominant PCR assay.  Population genetic summaries were calculated for a subset of individuals from four teosinte populations in central Mexico. Phenotypic data were also collected using seed from a single teosinte population where \emph{Hopscotch} was found segregating at high frequency.

@@

\textbf{Sampling and genotyping---}\lev{We sampled 1,110 individuals from 350 accessions (247 maize landraces, 17 \emph{mexicana} populations, and 86 \emph{parviglumis} populations}; ranging from 1-38 individuals per population, with an average of 11 individuals per population for \emph{parviglumis} and \emph{mexicana} and 2 individuals per landrace accession) and assessed the presence or absence of the \emph{Hopscotch} insertion (Appendix 1 and Appendix 2, See Supplemental Materials with the online version of this article). DNA was extracted from leaf tissue using a modified CTAB approach \citep{DoyleDoyle1990, Maloof1984}. We designed primers using PRIMER3 \citep{RozenSkaletsky2000} implemented in Geneious \citep{Kearse2012} to amplify the entire \emph{Hopscotch} element, as well as an internal primer allowing us to simultaneously check for possible PCR bias between presence and absence of the \emph{Hopscotch} insertion due to its large size ($\sim$5kb). Two PCRs were performed for each individual, one with primers flanking the \emph{Hopscotch} (HopF/HopR) and one with a flanking primer and an internal primer (HopF/HopIntR). Primer sequences are HopF, {\small 5'-TCGTTGATGCTTTGATGGATGG-3'}; 
HopR, {\small 5'-AACAGTATGATTTCATGGGACCG-3'}; and HopIntR, {\small  5'-CCTCCACCTCTCATGAGATCC-3'} (Appendix 3 and Appendix 4, See Supplemental Materials with the online version of this article). Homozygotes show a single band for absence of the element ($\sim$300bp) and two bands for presence of the element ($\sim$5kb, amplification of the entire element, and $\sim$1.1kb, amplification of part of the element), whereas heterozygotes show all three bands (Appendix 2, See Supplemental Materials with the online version of this article). Since we developed a PCR protocol for each allele, if only one PCR resolved well, we scored one allele for that individual rather than infer the diploid genotype. We used Phusion High Fidelity Enzyme (Thermo Fisher Scientific Inc., Waltham, Massachusetts, USA) and the following conditions for amplifications: 98$^{\circ}$C for 3 min, 30 cycles of 98$^{\circ}$C for 15 s, 65$^{\circ}$C for 30 s, and 72$^{\circ}$C for 3 min 30 s, with a final extension of 72$^{\circ}$C for 10 min. PCR products were visualized on a 1\% agarose gel and scored for presence/absence of the \emph{Hopscotch} based on band size.

@@

We examined evidence of introgression on chromosome 1 in these same four populations (EjuA, EjuB, MSA, SLO) using STRUCTURE \citep{Falush2003} and phased data from \citet{Pyhajarvi2013}, combined with the corresponding SNP data from a diverse panel of 282 maize lines \citep{Cook2012}. SNPs were anchored in a modified version of the IBM genetic map \citep{Gerke2013}. Since STRUCTURE does not account for LD due to physical linkage we created haplotype blocks using a custom Perl script from \citet[][code available at \url{http://dx.doi.org/10.6084/m9.figshare.1165577}]{Hufford2013}.  In maize, LD decays over an average distance of 5500bp \citep{Chia2012}; because LD decay is even more rapid in teosinte \citep{Chia2012} we used a conservative haplotype block size of 5kb.  We ran STRUCTURE at K=2 under the linkage model, with the assumption being that individuals fall into either a maize or teosinte cluster, performing 3 replicates with an MCMC burn-in of 10,000 steps and 50,000 steps post burn-in.

@@

\textbf{Genotyping---}The genotype at the \emph{Hopscotch} insertion was confirmed with two PCRs for \lev{837 individuals of the 1,100 screened}. Among the 247 maize landrace accessions genotyped, all but eight were homozygous for the presence of the insertion (Appendix 1 and Appendix 2, See Supplemental Materials with the online version of this article). Within our \emph{parviglumis} and \emph{mexicana} samples we found the \emph{Hopscotch} insertion segregating in 37 ($n=86$) and four ($n=17$) populations, respectively, and at highest frequency within populations in the states of Jalisco, Colima, and Michoac\'{a}n in central-western Mexico (Fig. \ref{Fig1Map}). Using our \emph{Hopscotch} genotyping, we calculated differentiation between populations (F\textsubscript{ST}) and subspecies (F\textsubscript{CT}) for populations in which we sampled eight or more individuals. We found that F\textsubscript{CT} = 0, and levels of F\textsubscript{ST} among populations within each subspecies (0.22) and among all populations (0.23) are similar to genome-wide estimates from previous studies (\citealt{Pyhajarvi2013}; Table \ref{Table1Fst}). Although we found large variation in \emph{Hopscotch} allele frequency among our populations, BayEnv analysis did not indicate a correlation between the \emph{Hopscotch} insertion and environmental variables (all Bayes Factors $<$ 1).

@@

%Figure 1 Map 
%--------------------------------------------
\begin{figure*}[!t]
\begin{center}
   \includegraphics[width=0.6\textwidth]{Fig1MapBalsas.pdf}
   \end{center}
    \caption{Map showing the frequency of the \emph{Hopscotch} allele in populations of \emph{parviglumis} where we sampled more than 6 individuals. Size of circles reflects number of alleles sampled. The Balsas River is shown, as the Balsas River Basin is believed to be the center of domestication of maize.} 
\label{Fig1Map}
\end{figure*}
%--------------------------------------------

% Figure 2. LD plots
%-------------------------------------------------------------------
\begin{figure*}[!t]
  \begin{center}
   \includegraphics[width=140mm]{Fig2LDPlots.pdf}
   \end{center}
    \caption{Linkage disequilibrium for SNPs in Mb 261-268 on chromosome 1. The yellow rectangle indicates the location of the \emph{Hopscotch} insertion and the green represents the \emph{tb1} ORF. A) Ejutla A; B) Ejutla B; C) La Mesa; D) San Lorenzo. The upper triangle above the black diagonal is colored based on the $r^{2}$ value between SNPs while the bottom triangle is colored based on p-value for the corresponding  $r^{2}$ value.}
\label{Fig2LD}
\end{figure*}
%-------------------------------------------------------------------

% Figure 3 STRUCTURE Plot 4 Pops
%-------------------------------------------------------------------
\begin{figure*}[!t]
  \begin{center}
   \includegraphics[width=90mm]{Fig3Structure.pdf}
     \end{center}
    \caption{STRUCTURE assignment to maize across a section of chromosome 1. The dotted lines mark the beginning of the sequenced region 66kb upstream (Region 2) and the end of the \emph{tb1} ORF.}
\label{Fig3Structure}
\end{figure*}
%-------------------------------------------------------------------

% Figure 4 Phenotyping BoxPlots
%-------------------------------------------------------------------
\begin{figure*}[!t]
  \begin{center}
   \includegraphics[width=90mm]{Figure4BoxplotPheno1.pdf}
   \end{center}
    \caption{Box-plots showing tillering index in greenhouse grow-outs for phenotyping. White indicates individuals homozygous for the \emph{Hopscotch}, light grey represents heterozygotes, and dark grey represents homozygotes for the teosinte (No \emph{Hopscotch}) allele. Within boxes, dark black lines represent the median, and the edges of the boxes are the first and third quartiles. Outliers are displayed as dots, the maximum value excluding outliers is shown with the top whisker, while the minimum excluding outliers is shown with the bottom whisker.}
\label{Fig4Boxplots}
\end{figure*}
%-------------------------------------------------------------------
